How to Not write Virtual Machines for Dynamic Languages

### 2007-05-27 talk/dyla2007/dyla.tex
A short dynamic compilers section.

A detailed description of this translation process is beyond the scope of the
present paper; it can be found in \cite{pypyvmconstruction}.  The actual Python
interpreter of PyPy and the results we achieved by translating it to C, LLVM
\cite{LLVM} and .NET are described in [XXX].  These results show that the
approach is practical and gives results whose performance is within the same
order of magnitude (within a factor of 2 and improving) of the hand-written,
well-tuned CPython, the C reference implementation.

@@

A more subtle example of separation of concerns is the way our generated
implementations can be integrated with a host OO VM.  As mentioned
above, an implementer deciding to directly target a specific OO VM needs
not only good knowledge of the OO VM in question and its object model --
he must fit the language into the imposed models.  Instead, in our
approach this task is done at two levels: in a first step, a stand-alone
interpreter is written -- which, if translated to a given OO VM, would
simply give an interpreter for the dynamic language which is unable to
communicate with the host VM.  Integration comes as a second step, and
occurs at a different level, by introducing mappings between the
relevant classes of the interpreter and the corresponding classes of the
OO VM.

@@

Although it is able to generate, among other things, a complete custom
VM for C-like environments, we found that the required effort that must
be put into the translation toolchain was still much lower than that of
writing a good-quality OO VM.  A reason seems to be that we could design
our translation toolchain specifically for our needs, i.e. a language
implementer's needs, instead of for general-purpose usage.  Of course,
the translation toolchain, once written, can also be reused to implement
other languages, and possibly tailored on a case-by-case basis to fit
the specific need of a language.  The process is incremental: we can add more
features as needed instead of starting from a maximal up-front design,
and gradually improve the quality of the tools, the garbage collectors,
the various optimizations, etc.

Let us expand on the topic of the garbage collector, which for C-like
envrionments is inserted into the generated VM by a transformation step.
We started by ignoring the issue and just using the conservative Boehm
\cite{Boehm} collector for C.  Later, we experimented with a range of
simple custom collectors - reference counting, mark-and-sweep, etc.
Ultimately, though, more advanced GCs will be needed to get the best
performance.  It seems that RPython, enhanced with support for direct
address manipulations, is a good language for writing GCs, so it would
be possible for a GC expert to write one for our translation framework.
However, this is not the only way to obtain good GCs: we will soon
investigate a more practical course of action, which is to reuse
existing GCs.  A good candidate is the GCs written in the Jikes RVM
\cite{JikesGC}.  As they are in Java, it should be relatively
straightforward to add a translation step that turns one of them into
RPython (or directly our RPython-level intermediate representation) and
integrate it with the rest of the program being translated.

In summary, developing a meta-programming translation toolchain requires
some work, but it can be done incrementally, it can reuse existing code,
and it gives a toolchain that is itself highly reusable and flexible in
nature.

>>>>

A detailed description of this translation process is beyond the scope of the
present paper; it can be found in \cite{pypyvmconstruction}.  The actual Python
interpreter of PyPy and the results we achieved by translating it to C, LLVM
\cite{LLVM} and .NET are described in [XXX].  These results show that the
approach is practical and gives results whose performance is within the same
order of magnitude (within a factor of 2 and improving) of the hand-written,
well-tuned CPython, the C reference implementation.  These figures do not
include the spectacular speed-ups obtained in some cases by the JIT compiler
described in section \ref{subsect:dynamic_compilers}.

@@

%A more subtle example of separation of concerns is the way our generated
%implementations can be integrated with a host OO VM.  As mentioned
%above, an implementer deciding to directly target a specific OO VM needs
%not only good knowledge of the OO VM in question and its object model --
%he must fit the language into the imposed models.  Instead, in our
%approach this task is done at two levels: in a first step, a stand-alone
%interpreter is written -- which, if translated to a given OO VM, would
%simply give an interpreter for the dynamic language which is unable to
%communicate with the host VM.  Integration comes as a second step, and
%occurs at a different level, by introducing mappings between the
%relevant classes of the interpreter and the corresponding classes of the
%OO VM.

@@

Although it is able to generate, among other things, a complete custom
VM for C-like environments, we found that the required effort that must
be put into the translation toolchain was still much lower than that of
writing a good-quality OO VM.  A reason is that a translation toolchain
operates in a more static way, which allows it to leverage good C
compilers.  It is self-supporting: pieces of the implementation can be
written in RPython as well and translated along with the rest of the
RPython source, and they can all be compiled and optimized by the C
compiler.  In order to write an OO VM in this style you need to start by
assuming an efficient dynamic compiler.

Of course, the translation toolchain, once written, can also be reused
to implement other languages, and possibly tailored on a case-by-case
basis to fit the specific needs of a language.  The process is
incremental: we can add more features as needed instead of starting from
a maximal up-front design, and gradually improve the quality of the
tools, the garbage collectors, the various optimizations, etc.

Writing a good garbage collector remains hard, though.  At least, it is
easy to experiment with various kind of GCs, so we started by just using
the conservative Boehm \cite{Boehm} collector for C and moved up to a
range of simple custom collectors -- reference counting, mark-and-sweep,
etc.  Ultimately, though, more advanced GCs will be needed to get the
best performance.  It seems that RPython, enhanced with support for
direct address manipulations, is a good language for writing GCs, so it
would be possible for a GC expert to write one for our translation
framework.  However, this is not the only way to obtain good GCs:
existing GCs can also be reused.  Good candidates are the GCs written in
the Jikes RVM \cite{JikesGC}.  As they are in Java, it should be
relatively straightforward to add a translation step that turns one of
them into RPython (or directly into our RPython-level intermediate
representation) and integrate it with the rest of the program being
translated.

In summary, developing a meta-programming translation toolchain requires
work, but it can be done incrementally, it can reuse existing code, and
it gives a toolchain that is itself highly reusable and flexible in
nature.

\subsection{Dynamic compilers}
\label{subsect:dynamic_compilers}

As mentioned above, the performance of the VMs generated by our
translation framework are quite acceptable -- e.g. the Python VM
generated via C code is much faster than Jython running on the best
JVMs.  Of course, the JIT compilers in these JVMs are essential to
achieve even this performance, which further proves the point that
writing good OO VMs -- especially ones meant to support dynamic
languages -- is a lot of work.

The deeper problem with the otherwise highly-tuned JIT compilers of the
OO VMs is that they are not a very good match for running dynamic
languages.  It might be possible to tune a general-purpose JIT compiler
enough, and write the dynamic language implementation accordingly, so
that most of the bookkeeping work involved in running the dynamic
language can be removed -- dispatching, boxing, unboxing...  However
this has not been demonstrated yet.

By far the fastest Python implementation, Psyco \cite{Psyco} contains a
hand-written language-specific dynamic compiler.  PyPy's translation
tool-chain is able to extend the generated VMs with an automatically
generated dynamic compiler similar to Psyco, derived from the
interpreter.  This is achieved by a pragmatic application of partial
evaluation techniques guided by a few hints added to the source of the
interpreter.  In other words, it is possible to produce a reasonably
good language-specific JIT compiler and insert it into a VM, alongside
with the necessary support code and the rest of the regular interpreter.

This result was one of the major goals and motivations for the whole
approach.  By construction, any code written in the dynamic language
runs correctly under the JIT.  Some very simple Python examples run more
than 100 times faster.  At the time of this writing this is still rather
experimental, and the techniques involved are well beyond the scope of
the present paper.  The reader is referred to \cite{D08.2} for more
information.
